Geospatial - Assignment 3

### 2025-02-13 hw3/hw3_MoritzPeistNoemiLucchiSimonVellin.Rmd
first commit for part 2

# Spatial data handling
library(sf) # Simple Features for spatial vector data
library(spData) # Spatial datasets
library(terra) # Raster data handling
library(exactextractr) # Precise spatial averaging

@@

>>>>

# Spatial data handling
library(sf) # Simple Features for spatial vector data
library(spData) # Spatial datasets
library(terra) # Raster data handling
library(exactextractr) # Precise spatial averaging
library(gdistance) # Compute minimum distances on raster surfaces

@@

To calculate the distances between the 10 top populated cities in Spain, and represent the roads within the continental Spain, we used data from the [Natural Earth website](https://www.naturalearthdata.com/downloads/10m-cultural-vectors/). Precisely, we loaded:
  - ne_10m_populated_places shapefile for the most populated cities 
  - ne_10m_roads for the streets

```{r}
# Load populated cities and roads
places <- st_read("/hw3/data/ne_10m_populated_places")
roads <- st_read("/hw3/data/ne_10m_roads")
```

## Data Wrangling
From the R package spData, we use the world dataset to extract the geometry of Spain, filtering only for continental Spain. Then, from the dataset containing the most populated cities, we extract only the cities in Spain, sort them in decreasing order of population, and retain only the top 10. Finally, from these top 10 cities, we extract Madrid and Vigo, which will be used in the plot.

```{r}
# Filter for continental Spain only 
spain <- world %>% 
  filter(name_long=='Spain') %>%
  st_cast("POLYGON") %>%                   
  slice_max(st_area(.))

# Get top 10 most populated Spanish cities
spain_top_10_cities <- places %>%
  filter(ADM0NAME == "Spain") %>%
  arrange(desc(POP_MAX)) %>%
  slice_head(n = 10)

# Filter Madrid and Vigo for the plot 
madrid_vigo <- spain_top_10_cities %>%
  filter(NAME %in% c("Madrid", "Vigo"))

```

## Methodological Approaches
First, we transform the CRS from EPSG:4326 to EPSG:3035, as the former uses latitude and longitude, which would lead to incorrect distance calculations in kilometers. Next, we intersect the road network with Spain’s boundaries to restrict the analysis to roads that are actually within Spain. We then create an empty raster over Spain, set a high resolution, and define the CRS as EPSG:3035 for the same reason mentioned earlier. After that, we generate the friction surface by assigning a value of 1 to cells that contain roads and 100 to cells that do not. This ensures that areas without roads have higher friction, influencing the computation of the shortest path between cities. Subsequently, we compute the transition matrix and, finally, calculate and print the distance matrix, which includes all pairwise distances between the top 10 most populated Spanish cities.

```{r}
# Transform to planar EPSG:3035 for better distance calculation in km 
spain_planar <- st_transform(spain, 3035)
roads_planar <- st_transform(roads, 3035)
cities_planar <- st_transform(spain_top_10_cities, 3035)

# Intersect roads with Spain's boundaries 
spain_roads <- st_intersection(roads_planar, spain_planar)

# Create and prepare a raster on the top of Spain 
raster <- rast(ext(spain_planar), resolution = 5000, crs = "EPSG:3035")  
roads_vect <- vect(spain_roads)
road_raster <- rasterize(roads_vect, raster, field = 1)

# Create friction surface: we assign 100 to cells without streets and 1 with cells 
# with streets. In this way we increase the friction for areas without streets, and so 
# streets will be chosen when computing the distances between places. 
road_values <- values(road_raster)
road_values[is.na(road_values)] <- 100
road_values[!is.na(road_values)] <- 1
values(road_raster) <- road_values

# Create transition matrix for distance calculation
tr_matrix <- transition(raster(road_raster), 
                        transitionFunction = mean, 
                        directions = 8)

# Calculate distances between all city pairs
n_cities <- nrow(cities_planar)
dist_matrix <- matrix(0, n_cities, n_cities)
rownames(dist_matrix) <- cities_planar$NAME
colnames(dist_matrix) <- cities_planar$NAME

for(i in 1:n_cities) {
  for(j in 1:n_cities) {
    if(i < j) {  
      path <- shortestPath(tr_matrix, 
                           origin = as_Spatial(cities_planar[i,]), 
                           goal = as_Spatial(cities_planar[j,]), 
                           output = "SpatialLines")
      path_sf <- st_as_sf(path)
      st_crs(path_sf) <- st_crs(cities_planar)
      # Calculate distance in kilometers
      dist <- as.numeric(st_length(path_sf)) / 1000
      dist_matrix[i,j] <- dist
      dist_matrix[j,i] <- dist}}}

# Print the distances
print("Distances between cities (km):")
print(round(dist_matrix, 1))
```

## Plotting 
After these manipulations, we can plot continental Spain with its roads, highlighting Madrid and Vigo, as shown in the reference plot.

```{r}
# Plotting
ggplot() +
  geom_sf(data = st_transform(spain, 4326), fill = "lightgray") +
  geom_sf(data = st_transform(spain_roads, 4326), color = "black", size = 0.3) +
  geom_sf(data = st_transform(spain_top_10_cities, 4326), size = 1, color = "black") +
  geom_sf(data = st_transform(madrid_vigo %>% filter(NAME == "Madrid"), 4326), 
          aes(shape = "Madrid", color = "Madrid"), size = 3) +
  geom_sf(data = st_transform(madrid_vigo %>% filter(NAME == "Vigo"), 4326), 
          aes(shape = "Vigo", color = "Vigo"), size = 3) +
  scale_color_manual(name = "Cities",
                     values = c("Madrid" = "red", "Vigo" = "turquoise")) +
  scale_shape_manual(name = "Cities",
                     values = c("Madrid" = 16, "Vigo" = 17)) 
```
